Importance of depth refugia for reef resilience and intervention on the Great Barrier Reef under future climate change

### 2025-10-08 paper/paper_draft.qmd
Updates with tracked changes for methods section

The spatial extent of reefs represented in this investigation follows reef definitions
available via the Great Barrier Reef Marine Park Authority (GBRMPA), defining 3806 reefs
across the GBR [@GBRMPA2022] (@fig-region-map). Estimates of coral carrying capacity for
each reef were taken from prior analyses [@bozec2022].

@@

An exploratory scenario modelling approach was adopted for this investigation. The method
involves exploring a wide range of potential futures and model parameterisations under
different assumptions of environmental (climatic) and ecological conditions. Exploratory
modelling approaches have gained favour in recent years to explore the implications of deep
uncertainty; where there is little or no agreement on the key drivers of a system, the
extent of the relationship, how best to represent them or their probability distributions
[@conradt2024; @lempert2013]. The approach embraces uncertainty as an inherent
characteristic of complex environmental systems [@lempert2010; @lempert2013] in an attempt
to address limitations of calibrated ecosystem models that use a singular set of parameter
values [@lubianabotelho2025]. Exploratory scenario modelling is a core methodology
underpinning the development of robust and adaptive management strategies [@haasnoot2013].

All data preparation, modelling and analyses were performed using the Julia programming
language (The Julia Project; https://julialang.org/; [@bezanson2017]), with visualisations
created using the Makie plotting ecosystem [@danisch2021]. Code used to process DHW data and
perform analyses involved in this investigation can be found at Grier and Iwanaga
[-@grier2025].

@@

The coral larval connectivity model used in this study comprises 3D hourly velocities from
the 1 km grid resolution hydrodynamic model (GBR-1 version 2.0) [@steven2019], as input to
OceanParcels [@delandmeter2019], a particle tracking model. The connectivity model was used
to simulate larval dispersal and settlement of broadcast-spawning Acropora corals in the
GBR, following annual mass spawning events in 2015, 2016, 2018, 2019 and split-spawning
events in 2017 and 2020. Split spawning occurs when the annual reproductive event of a coral
population is split over two consecutive months. Coral mass spawning usually occurs 4-6 days
after the full moon during the spring-summer transition in the GBR with variations across
species and locations. The spawning days identified are: 2015 - November 30, December 1 and
2; 2016 - November 18, 19 and 20; 2017 - November 8, 9 and 10, December 8, 9 and 10; 2018 –
November 18, 19, 20; 2019 – November 16, 17, 18; 2020 – November 5, 6 and 7, December 4, 5,
6.

@@

Larval transition matrices indicate the probability of larval transfer between pairs of
reefs in the GBR, herein referred to as connectivity strengths. Indicative connectivity
strengths are implemented in ADRIA-CoralBlox as the mean of matrices for each location
(i.e., mean of all 24 spawning days). While the connectivity strengths do not vary
temporally within each ADRIA-CoralBlox simulation, spatial variability enables
investigations into the influence of connectivity strength and its role in driving recovery
processes.

To assess the impact of reef connectivity on modelled outcomes, connectivity was summarised
by incoming connectivity (weighted incoming connectivity), outgoing connectivity (outgoing
connection strength), the ratio of source-to-sink connectivity, and larval retention
probability for each reef. Weighted incoming connectivity was calculated by summing the
product of each connection strength incoming to a target reef (excluding retained larvae
from the target reef) and the mean timeseries absolute coral cover (km^2^) of each matching
source reef. This metric provides an approximation of overall larval flow into a reef, with
high values indicating the target reef is supplied high volumes of larvae via strong
connections and low values indicating the target reef has weak (or no) incoming connections
or is supplied by small source populations. A limitation of weighted incoming connectivity
is that it does not incorporate the proportion of the source reef's population that is
reproducing and producing larvae, instead it takes total coral cover area as an indicator of
larval productivity. Outgoing connection strength was calculated as the sum of all outgoing
transition probabilities from a target reef. Outgoing connection strength has a theoretical
maximum of one (all larvae reach settlement on sink reefs), however the large size of the
GBR combined with larval mortality result in much lower outgoing connection strength values.
The ratio of source-to-sink connectivity ($O:I$) for a target reef represents the strength
of the reef's outgoing larval flow relative to it's incoming larval flow. This is calculated
as: $O:I = \frac{S_O/N_O}{S_I/N_I}$. Where $S$ represents the sum of connectivity
probabilities, $N$ represents the number of connections, $O$ denotes outgoing connectivity
from the target reef and $I$ denotes incoming connectivity to the target reef. If a reef has
no incoming connections, the $O:I$ equals the $S_O/N_O$ value. Incoming connections in this
calculation, do not directly represent the larval supply gained as the breeding populations
from which connections are coming from are not represented. A source-to-sink ratio greater
than 1 (a value greater than 0 on a log scale) indicates a reef has greater outgoing than
incoming larval flow. Reef larval retention probability indicates the proportion of spawned
larvae that settle on their natal reefs.

@@

The coral demographic representation and associated model implementation are further
detailed in the supplementary material. Of note is that the representation of bleaching
mortalities followed Bozec et al., [-@bozec2022] albeit with a scaling coefficient to
incorporate the amelioration effect that depth may have, following Baird et al.,
[@baird2018]. This depth-scaling reduces the proportion of coral colonies impacted by high
heat stress as depth increases. The impacts of tropical cyclones, wave activity, Crown of
Thorns Starfish or coral disease were not included in this investigation in order to focus
on assessing reef resilience under possible future warming scenarios.

ADRIA-CoralBlox was run for 4096 scenarios for each GCM (20,480 total model runs). Here, a
scenario represents a unique combination of assumed environmental (namely one of the five
DHW trajectories) and ecological factors including coral growth and mortality rates, and
coral thermal tolerances. Scenarios were generated using the low-discrepancy Sobol' sampling
method [@sobol1967; @sobol1993] with Owen Scrambling [@owen1995], a randomised Quasi Monte
Carlo approach which balances scenario exploration with improved convergence properties
compared to pseudo-random sampling. A detailed comparative analysis of sampling methods can
be found in [@kucherenko2015]. For scenario assessment, coral cover time series for each
model run were scaled to represent cover as a percentage of the total estimated reef area.
The scenario-median time series for each reef was used for further analyses. Time series of
coral cover were then clustered based on their temporal behaviour. Assessment metrics used
in analyses can be found in @stbl-metrics.

@@

Reefs were clustered based on the temporal behaviour of their median coral cover time
series. Put simply, reefs that have similar trends and trajectories are more likely to be
clustered together. Reefs are clustered into Low, Medium, or High, indicating their relative
ability to maintain coral cover over time. Reefs in each time series cluster were compared
to identify characteristics that differ between Low, Medium and High clusters. These
characteristics include reef median depth, mean time series DHW, weighted incoming
connectivity, outgoing connection strength, ratio of source-to-sink larval connectivity,
larval retention probability and reef carrying capacity (coral habitable area).

>>>>

The spatial extent of reefs represented in this investigation follows reef definitions
available via the Great Barrier Reef Marine Park Authority (GBRMPA). Reef definitions outline 3806 reefs
across the GBR [@GBRMPA2022] (@fig-region-map). Prior analyses by Bozec et al. [-@bozec2022] provided estimates of coral carrying capacity for each reef that were adopted by this study.

@@

An exploratory scenario modelling approach was adopted for this investigation. The method
involves exploring a wide range of potential futures and model parameterisations under
different assumptions of climatic and ecological conditions. Exploratory
modelling approaches have gained favour in recent years to explore the implications of deep
uncertainty; where there is little or no agreement on the key drivers of a system, the
extent of the relationship, how best to represent them or their probability distributions
[@conradt2024; @lempert2013]. The approach embraces uncertainty as an inherent
characteristic of complex environmental systems [@lempert2010; @lempert2013] 
to address limitations of calibrated ecosystem models that use a singular set of parameter
values [@lubianabotelho2025]. Exploratory scenario modelling is a core methodology
underpinning the development of robust and adaptive management strategies [@haasnoot2013].

All data preparation, modelling and analyses were performed using the Julia programming
language [@bezanson2017], with visualisations
created using the Makie plotting ecosystem [@danisch2021]. Code used to process DHW data and
perform analyses involved in this investigation can be found at Grier and Iwanaga
[-@grier2025].

@@

The coral larval connectivity model used in this study comprises 3D hourly velocities from
the 1 km grid resolution hydrodynamic model [GBR-1 version 2.0; @steven2019], as input to
OceanParcels [@delandmeter2019], a particle tracking model. The connectivity model was used
to simulate larval dispersal and settlement of broadcast-spawning Acropora corals in the
GBR, following annual mass spawning events in 2015, 2016, 2018, 2019 and split-spawning
events in 2017 and 2020. Split spawning occurs when the annual reproductive event of a coral
population is split over two consecutive months. Coral mass spawning usually occurs 4-6 days
after the full moon during the spring-summer transition in the GBR with variations across
species and locations. The spawning days identified are: 2015 - November 30, December 1 and
2; 2016 - November 18, 19 and 20; 2017 - November 8, 9 and 10, December 8, 9 and 10; 2018 –
November 18, 19, 20; 2019 – November 16, 17, 18; 2020 – November 5, 6 and 7, December 4, 5,
6.

@@

Larval transition matrices indicate the probability of larval transfer between pairs of
reefs in the GBR, herein referred to as connectivity strengths. Indicative connectivity
strengths are implemented in ADRIA-CoralBlox (described in the next section) as the mean of matrices for each location
(i.e., mean of all 24 spawning days). While the connectivity strengths do not vary
temporally within each ADRIA-CoralBlox simulation, spatial variability enables
investigations into the influence of connectivity strength and its role in driving recovery
processes.

To assess the impact of reef connectivity on modelled outcomes, connectivity was summarised
by incoming connectivity (weighted incoming connectivity), outgoing connectivity (outgoing
connection strength), the ratio of source-to-sink connectivity, and larval retention
probability for each reef. Weighted incoming connectivity was calculated by summing the
product of each connection strength incoming to a target reef (excluding retained larvae
from the target reef) and the mean timeseries absolute coral cover (km^2^) of each matching
source reef. This metric provides an approximation of overall larval flow into a reef, with
high values indicating the target reef is supplied high volumes of larvae via strong
connections and low values indicating the target reef has weak (or no) incoming connections
or is supplied by small source populations. A limitation of weighted incoming connectivity
is that it does not incorporate the proportion of the source reef's population that is
reproducing and producing larvae, instead it takes total coral cover area as an indicator of
larval productivity.

Outgoing connection strength was calculated as the sum of all outgoing
transition probabilities from a target reef. Outgoing connection strength has a theoretical
maximum of one (all larvae reach settlement on sink reefs), however the large size of the
GBR combined with larval mortality result in much lower outgoing connection strength values.
The ratio of source-to-sink connectivity ($O:I$) for a target reef represents the strength
of the reef's outgoing larval flow relative to it's incoming larval flow. This is calculated
as: $O:I = \frac{S_O/N_O}{S_I/N_I}$. Where $S$ represents the sum of connectivity
probabilities, $N$ represents the number of connections, $O$ denotes outgoing connectivity
from the target reef and $I$ denotes incoming connectivity to the target reef. If a reef has
no incoming connections, the $O:I$ equals the $S_O/N_O$ value. Incoming connections in this
calculation, do not directly represent the larval supply gained as the breeding populations
from which connections are coming from are not represented. A source-to-sink ratio greater
than 1 (a value greater than 0 on a log scale) indicates a reef has greater outgoing than
incoming larval flow. Reef larval retention probability indicates the proportion of spawned
larvae that settle on their natal reefs.

@@

The coral demographic representation and associated model implementation are further
detailed in the supplementary material. Of note is that the representation of bleaching
mortalities followed Bozec et al., [-@bozec2022] and  albeit with a scaling coefficient to
incorporate the amelioration effect that depth may have, following Baird et al.,
[-@baird2018]. This depth-scaling reduces the proportion of coral colonies impacted by high
heat stress as depth increases. The impacts of tropical cyclones, wave activity, Crown of
Thorns Starfish or coral disease were not included in this investigation in order to focus
on assessing reef resilience under possible future warming scenarios.

ADRIA-CoralBlox was run for 4096 scenarios for each GCM (20,480 total model runs). Here, a
scenario represents a unique combination of assumed environmental (namely one of the five
DHW trajectories) and ecological factors including coral growth and mortality rates, and
coral thermal tolerances. Scenarios were generated using the low-discrepancy Sobol' sampling
method [@sobol1967; @sobol1993] with Owen Scrambling [@owen1995], a randomised Quasi Monte
Carlo approach which balances scenario exploration with improved convergence properties
compared to pseudo-random sampling. A detailed comparative analysis of sampling methods can
be found in [@kucherenko2015]. For scenario assessment, coral cover time series for each
model run were scaled to represent cover as a percentage of the total estimated reef area.
The scenario-median time series for each reef was used for further analyses. Time series of
coral cover were then clustered based on their behaviour. Assessment metrics used
in analyses can be found in @stbl-metrics.

@@

Reefs were clustered based on the behaviour of their median coral cover time
series. Put simply, reefs that have similar trends and trajectories are more likely to be
clustered together. Reefs are clustered into Low, Medium, or High, indicating their relative
ability to maintain coral cover over time. Reefs in each time series cluster were compared
to identify characteristics that differ between Low, Medium and High clusters. These
characteristics include reef median depth, mean time series DHW, weighted incoming
connectivity, outgoing connection strength, ratio of source-to-sink larval connectivity,
larval retention probability and reef carrying capacity (coral habitable area).
